---
title: "Music Data Analysis Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Music Data Analysis Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This vignette shows how to use the `musicDataAnalysis` package. The package helps clean, summarize, and visualize the Top Spotify Songs 2023 dataset. The goal is to understand patterns in music popularity using streams and audio features such as energy, danceability, and bpm.

## Load the package and sample data

```{r}
library(musicDataAnalysis)

spotify_sample <- load_spotify_sample()
head(spotify_sample)
```

The included sample data is already cleaned. It has track names, artist names, streams, energy, danceability, and bpm.

## Summary statistics

```{r}
summarize_music(spotify_sample)
```

This output gives the average number of streams and the correlations between streams and the audio features.

```{r}
summarize_features(spotify_sample)
```

This output gives basic statistics for energy, danceability, and bpm.

## Artist analysis

```{r}
top_artists(spotify_sample, n = 10)
```

This shows the artists with the highest total streams in the cleaned data.

```{r, warning=FALSE}
artist_summary(spotify_sample, "Taylor Swift")
```

This gives a short summary for one selected artist.

## Visualizations

```{r, fig.width = 7, fig.height = 5}
plots <- plot_music(spotify_sample)
plots$scatter_energy_streams
```

The scatterplot helps show the relationship between energy and streams.

```{r, fig.width = 7, fig.height = 5}
plot_feature_distribution(spotify_sample, "bpm")
```

This plot shows the distribution of bpm across songs.

## Conclusion

This package gives users a simple way to clean Spotify data, calculate basic summaries, and create useful plots. These tools make it easier to explore what characteristics appear in popular songs.
